### 2021-07-09 rad-icfp-2021/abstract.tex
Typos and fixes.

\begin{abstract}
    Reverse-mode differentiation is used for optimization, but it introduces references which break the purity of the underlying program, making them notoriously harder to optimize.
    We present a reverse-mode differentiation on a purely functional language with array operations.  
    It is the first one to deliver a provably efficient, purely functional, and denotationally correct reverse-mode differentiation.
    We show that our transformation is semantically correct and verifies the cheap gradient principle. 
    Inspired by PROPs and compilation to categories, we introduce a novel intermediate representation that we call `unary form'.
    Our reverse-mode transformation is factored as a compilation scheme through this intermediate representation.
    We obtain provably efficient gradients by performing general partial evaluation optimizations after our reverse-mode transformation, as opposed to manually derived ones.
    For simple first-order programs, the obtained output programs resemble static-single-assignment (SSA) code. 
    We emphasize the modularity of our approach and show how our language can easily be enriched with more optimized primitives, as required for some speedups in practice.
\end{abstract}

>>>>

\begin{abstract}
    Reverse-mode differentiation is used for optimization, but it introduces references, which break the purity of the underlying programs, making them notoriously harder to optimize.
    We present a reverse-mode differentiation on a purely functional language with array operations.  
    It is the first one to deliver a provably efficient, purely functional, and denotationally correct reverse-mode differentiation.
    We show that our transformation is semantically correct and verifies the cheap gradient principle. 
    Inspired by PROPs and compilation to categories, we introduce a novel intermediate representation that we call `unary form'.
    Our reverse-mode transformation is factored as a compilation scheme through this intermediate representation.
    We obtain provably efficient gradients by performing general partial evaluation optimizations after our reverse-mode transformation, as opposed to manually derived ones.
    For simple first-order programs, the obtained output programs resemble static-single-assignment (SSA) code. 
    We emphasize the modularity of our approach and show how our language can easily be enriched with more optimized primitives, as required for some speedups in practice.
\end{abstract}

### 2021-07-09 rad-icfp-2021/intro.tex
Typos and fixes.

% \subsection{Motivation}
Deep learning is moving towards increasingly sophisticated optimization objectives that employ tensors and operations on tensors.
Reverse-mode Automatic Differentiation (AD) is a technique to automatically compute the gradient of objective functions of form $\RR^n\to\RR$.
Such functions appear a lot in practice: for instance, as loss functions in machine learning.

In order to reach the efficiency of the usual imperative version of reverse-mode, the transformations even in functional languages~\cite{pearlmutter2008reverse} resort to references.
The lack of purity in reverse-mode is known to make it significantly harder to optimize and parallelize. 
None of the current implementations of reverse-mode in a functional setting~\cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g.~\cite{xla}.
As a result, to optimize for efficiency, a hand-crafted reverse-derivative must be given, for every important non-trivial operation.
In other words, abstracting away from imperative code is still a hurdle that functional implementations need to overcome.

@@

In this paper, we define a purely functional (without references or control mechanisms such as state monads), denotationally correct, and provably efficient reverse-mode AD. 
To do so, we define the Unary Normal Form (UNF) representation inspired by PROPs \cite{hackney2015category} and compilation to categories \cite{elliott2017compiling}.
We can easily define and prove correctness of reverse-mode on this representation. 
The whole reverse-mode transformation is obtained by compiling the language to this intermediate representation, applying the simpler reverse-mode transformation, and compiling again to the original language.
After standard optimizations, the output program looks like SSA~\cite{cytron1989efficient} or ANF~\cite{sabry1993reasoning}, which leads to more efficient implementations.

@@

We introduce the general idea of efficient reverse-mode in a functional setting the next examples.
% We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.

>>>>

% \subsection{Motivation}
Deep learning is moving towards increasingly sophisticated optimization objectives that employ tensors and operations on tensors.
Reverse-mode Automatic Differentiation (AD) is a technique to automatically compute the gradient of objective functions of the form $\RR^n\to\RR$.
Such functions appear a lot in practice: for instance, as loss functions in machine learning.

In order to reach the efficiency of the usual imperative version of reverse-mode, the transformations, even in functional languages~\cite{pearlmutter2008reverse}, resort to references.
The lack of purity in reverse-mode is known to make it significantly harder to optimize and parallelize. 
None of the current implementations of reverse-mode in a functional setting~\cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g.~\cite{xla}.
As a result, to optimize for efficiency, a hand-crafted reverse-derivative must be given, for every important non-trivial operation.
In other words, abstracting away from imperative code is still a hurdle that functional implementations need to overcome.

@@

In this paper, we define a purely functional (without references or control mechanisms such as state monads), denotationally correct, and provably efficient reverse-mode AD. 
To do so, we define the Unary Normal Form (UNF) representation inspired by PROPs~\cite{maclane1965categorical} %\cite{hackney2015category} 
and compilation to categories~\cite{elliott2017compiling}.
We can easily define and prove correctness of reverse-mode on this representation. 
The whole reverse-mode transformation is obtained by compiling the language to this Intermediate Representation (IR), applying the simpler reverse-mode transformation, and compiling again to the original language.
After standard optimizations, the output program looks like SSA~\cite{cytron1989efficient} or ANF~\cite{sabry1993reasoning}, which leads to more efficient implementations.

@@

We introduce the general idea of efficient reverse-mode in a functional setting through the following examples.
% We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.
